My wonderful paper

### 2024-12-04 _manuscript/_tex/index.tex
refine introduction

\begin{itemize}
\tightlist
\item
  emphasize trustworthy data science since it is the theme of the
  special issue
\end{itemize}

In a data analysis, an experienced researcher can often quickly assess
whether a result meets their expectation, while others may be unsure of
what to anticipate from the analysis or how to recognize when results
diverge form expected norms. (These expectations are often based on
prior knowledge, domain expertise, or common sense and they can often be
framed into a Boolean question that can be answered by a unit test. For
example, in a linear regression model, one may expect the p-value of the
coefficient to be less than 0.05. On top of the expectation on the final
results, researchers may also have expectations on the intermediate
steps of the analysis process. For example, one may expect the data
doesn't contain outliers, etc. These expectations are often checked
early on to prevent an unexpected propagate through the analysis
process.)

This expectation-setting process is often implicit and rarely documented
or discussed in publications. (There are most publication requires open
source code for publication, however, it is no instruction or guidance
on checking \ldots). Yet, making them explicit is crucial for 1) helping
junior researchers interpret results and diagnose the unexpected, 2)
providing checkpoints that support independent replication or
application of methods to new data, 3) aligning assumption across
researchers from different backgrounds.

In this paper, we formalize these informal checks into an \emph{analysis
plan} comprising the analysis steps along with the associated unit
tests. This formalization allows us to examine assumptions made about
the data during analysis and to compare different unit tests in
diagnosing unexpected outcomes. We then introduce a procedure to assess
the quality of these unit tests based on their accuracy and redundancy
across simulated version of the original data. The accuracy is assessed
using binary classification metrics -- precision and recall -- while
redundancy is measured using mutual information. The workflow provides
numerical guarantee of the consistency of the expectation and the data
for the analysis plan. This means that given the assumption of the data
generating mechanism, the analysis will provide the outcome we expect.
In data analysis, analysts are not necessarily collecting the data and
they can only make reasonable assumption about the quality of the data.
This method we propose provides a numerical guarantee, based on
simulation, that the analysis will produce the outcome we expect, given
the assumption of the data generating mechanism. This provides
transparency of the analysis and build trustworthy data analysis.

The rest of the paper is organized as follows: Section~\ref{sec-plan}
describes the concept of analysis plan in detail.
Section~\ref{sec-examples} provides examples of analysis plan {[}more
details{]}. (need another section here or before examples?)
Section~\ref{sec-conclusion} concludes the paper.

\section{Literature review}\label{literature-review}

@@

@@

\section{Method}\label{method}

@@

\section{Discussion}\label{discussion}

>>>>

\begin{itemize}
\tightlist
\item
  TODO: emphasize trustworthy data science since it is the theme of the
  special issue
\end{itemize}

In data analysis, experienced researchers often rely on their prior
knowledge or domain expertise to quickly assess whether results align
with their expectations. When outcomes deviate, they can quickly iterate
through the analysis cycle to identify potential issues, refine methods,
or revise assumptions. However, this iterative process they employ is
rarely explicit, documented, or discussed in publications. As a result,
readers are typically presented with the final outcomes of the analysis
cycle where the results and expectations are aligned -- achieved either
by refining the analysis or updating the expectations based on
statistical evidence. These missing pieces of information provides
little guidance for diagnosing issues in the analysis when the same
methodology is applied to a new dataset that produces different
outcomes. Similarly, when researchers with different background
knowledge it becomes unclear whether variations in results arise from
differences in the assumptions or in the analysis itself.

In this paper, we frame these expectations as checks within an
\emph{analysis plan}, which allows us to examine these implicit
assumptions made about the data during analysis and to compare different
checks for diagnosing unexpected outcomes. We then introduce a procedure
to evaluate the quality of these checks, either individually or in
combination through logic regression, based on their accuracy and
redundancy across simulation of the original data. Accuracy is assessed
using binary classification metrics, precision and recall, while
redundancy is measured using mutual information. The proposed workflow
offers a numerical guarantee, based on simulation, that the analysis
will produce the expected results, assuming the data generating
mechanism holds. {[}something about trustworthy data science{]}

The rest of the paper is organized as follows:
Section~\ref{sec-lit-review} reviews the concept of diagnosing
unexpected outcomes and data quality checks. Section~\ref{sec-plan}
introduces the concept of analysis plan, illustrated with a toy example
on fitness step count. Section~\ref{sec-method} describes the procedure
for selecting optimal check combinations to predict unexpected outcomes.
Section~\ref{sec-examples} applies this selection procedure to a linear
regression example examining the effect of PM10 on mortality.
Section~\ref{sec-discussion} discusses a few key considerations and
Section~\ref{sec-conclusion} concludes the paper.

\section{Literature review}\label{sec-lit-review}

@@

Yet, making them explicit is crucial for 1) helping junior researchers
interpret results and diagnose the unexpected, 2) providing checkpoints
that support independent replication or application of methods to new
data, 3) aligning assumption across researchers from different
backgrounds.

@@

\section{Method}\label{sec-method}

@@

\section{Discussion}\label{sec-discussion}

### 2024-12-04 index.qmd
refine introduction

-   emphasize trustworthy data science since it is the theme of the special issue

In a data analysis, an experienced researcher can often quickly assess whether a result meets their expectation, while others may be unsure of what to anticipate from the analysis or how to recognize when results diverge form expected norms. (These expectations are often based on prior knowledge, domain expertise, or common sense and they can often be framed into a Boolean question that can be answered by a unit test. For example, in a linear regression model, one may expect the p-value of the coefficient to be less than 0.05. On top of the expectation on the final results, researchers may also have expectations on the intermediate steps of the analysis process. For example, one may expect the data doesn't contain outliers, etc. These expectations are often checked early on to prevent an unexpected propagate through the analysis process.)

This expectation-setting process is often implicit and rarely documented or discussed in publications. (There are most publication requires open source code for publication, however, it is no instruction or guidance on checking ...). Yet, making them explicit is crucial for 1) helping junior researchers interpret results and diagnose the unexpected, 2) providing checkpoints that support independent replication or application of methods to new data, 3) aligning assumption across researchers from different backgrounds.

In this paper, we formalize these informal checks into an *analysis plan* comprising the analysis steps along with the associated unit tests. This formalization allows us to examine assumptions made about the data during analysis and to compare different unit tests in diagnosing unexpected outcomes. We then introduce a procedure to assess the quality of these unit tests based on their accuracy and redundancy across simulated version of the original data. The accuracy is assessed using binary classification metrics -- precision and recall -- while redundancy is measured using mutual information. The workflow provides numerical guarantee of the consistency of the expectation and the data for the analysis plan. This means that given the assumption of the data generating mechanism, the analysis will provide the outcome we expect. In data analysis, analysts are not necessarily collecting the data and they can only make reasonable assumption about the quality of the data. This method we propose provides a numerical guarantee, based on simulation, that the analysis will produce the outcome we expect, given the assumption of the data generating mechanism. This provides transparency of the analysis and build trustworthy data analysis.

The rest of the paper is organized as follows: @sec-plan describes the concept of analysis plan in detail. @sec-examples provides examples of analysis plan \[more details\]. (need another section here or before examples?) @sec-conclusion concludes the paper.

# Literature review

@@

# Method

@@

# Discussion

>>>>

-   TODO: emphasize trustworthy data science since it is the theme of the special issue

In data analysis, experienced researchers often rely on their prior knowledge or domain expertise to quickly assess whether results align with their expectations. When outcomes deviate, they can quickly iterate through the analysis cycle to identify potential issues, refine methods, or revise assumptions. However, this iterative process they employ is rarely explicit, documented, or discussed in publications. As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned -- achieved either by refining the analysis or updating the expectations based on statistical evidence. These missing pieces of information provides little guidance for diagnosing issues in the analysis when the same methodology is applied to a new dataset that produces different outcomes. Similarly, when researchers with different background knowledge it becomes unclear whether variations in results arise from differences in the assumptions or in the analysis itself.

In this paper, we frame these expectations as checks within an *analysis plan*, which allows us to examine these implicit assumptions made about the data during analysis and to compare different checks for diagnosing unexpected outcomes. We then introduce a procedure to evaluate the quality of these checks, either individually or in combination through logic regression, based on their accuracy and redundancy across simulation of the original data. Accuracy is assessed using binary classification metrics, precision and recall, while redundancy is measured using mutual information. The proposed workflow offers a numerical guarantee, based on simulation, that the analysis will produce the expected results, assuming the data generating mechanism holds. [something about trustworthy data science]

The rest of the paper is organized as follows: @sec-lit-review reviews the concept of diagnosing unexpected outcomes and data quality checks. @sec-plan introduces the concept of analysis plan, illustrated with a toy example on fitness step count. @sec-method describes the procedure for selecting optimal check combinations to predict unexpected outcomes. @sec-examples applies this selection procedure to a linear regression example examining the effect of PM10 on mortality. @sec-discussion discusses a few key considerations and @sec-conclusion concludes the paper.

# Literature review {#sec-lit-review}

@@

# Method {#sec-method}

@@

# Discussion {#sec-discussion}
